---
title: "Convergence of Evidence: land degrdation in Europe"
author: "M. Weynants"
date: "18/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(terra)
library(sf)
library(dplyr)
library(tidyr)
library(ggplot2)
wd <- getwd()
if (!dir.exists("./temp")){dir.create("./temp")}
```
## Define study area and zones
The outputs of the convergence of evidence are summarised at administrative (or other) level.
```{r zone_load}
# get official NUTS areas
if (!"eurostat" %in% installed.packages()[,1]){
  install.packages("eurostat")
}
library(eurostat)
reg_area3 <- get_eurostat("reg_area3") %>%
  filter(time == as.Date("2016-01-01")) %>%
  dplyr::select(geo, landuse, values) %>%
  # landuse: TSA (total surface area) = TOTAL
  #          TLA (total land area) = L0008
  pivot_wider(id_cols = geo, names_from = landuse, values_from = values) %>%
  rename(TSA = TOTAL, TLA = L0008)

if (!file.exists("./temp/zones_eu27.geojson")) {
# load NUTS level 3 polygons in ETRS89 from EUROSTAT gisco API
# Map Scale = Raster resolution (in meters) * 2 * 1000
# 2e6 doesn't exist so take 1:1000000
url_zones <- "https://gisco-services.ec.europa.eu/distribution/v2/nuts/geojson/NUTS_RG_01M_2021_3035_LEVL_3.geojson"
zones <- geojsonsf::geojson_sf(url_zones)
# cite as © EuroGeographics for the administrative boundaries
# crs not properly assigned
st_crs(zones) <- st_crs(3035)

# restrict to EU27
clip <- read_sf("../EU_dissolve/EU_dissolve.shp") %>%
  st_transform(crs = 3035)
zones_eu27 <- zones %>% 
  # EU27
  dplyr::filter(CNTR_CODE %in% c("BE", "BG", "CZ", "DK", "DE", "EE", "IE",
                                 "EL", "ES", "FR", "HR", "IT", "CY", "LV",
                                 "LT", "LU", "HU", "MT", "NL", "AT", "PL",
                                 "PT", "RO", "SI", "SK", "FI", "SE")) %>%
  # remove oversea territories !!! Cyprus is missing!!!
  sf::st_intersection(clip)%>%
  # add back missing CY
  add_row(zones %>% filter(CNTR_CODE == "CY")) %>%
  # add ID
  mutate(ID = row_number()) %>%
  # add area from eurostat
  dplyr::left_join(reg_area3, by = c("NUTS_ID" = "geo")) 
# export clip
st_write(zones_eu27, dsn = "./temp/zones_eu27.geojson")
# reload file as SpatVect
zones_eu27 <- vect(zones_eu27)
rm(zones, clip)
} else {
  zones_eu27 <- vect(geojsonsf::geojson_sf("./temp/zones_eu27.geojson"))
}

```

## Load CoE output data
Assuming all issues have been processed, so that they :
- have the same equal area geographical coordinate system
- have the same extent
- are all dummy variables with values of 0 and 1
- have the same definition for nodata

```{r CoE_load}
coe_files <- list.files( 
  path = "../CoE_Europe_OUTPUT/3035_1221",
  pattern = "*.tif$",
  full.names = TRUE)

# get band order from vrt
band_order <-  c(
  "coe_europe_december6" = "Sum of issues",
  "lpd_coe1000" = "Land Productivity Dynamics Change",
  "TreeLoss_1000_3035_5" = "Tree loss",
  "acidification" = "Soil acidification",
  "builtup" = "Built-up",
  "bws" = "Water stress",
  "eutrof" = "Eutrophication",
  "fires" = "Fires",
  "groundwaterdecline" = "Groundwater decline",
  "hhi" = "Household income",
  "irrigation" = "Irrigation under water stress",
  "nitrogen" = "Nitrogen",
  "pop" = "Population",
  "rusle" = "Water erosion",
  "wind" = "Wind erosion"
)
# coe_files <- list.files( 
#   path = "../CoE_Europe_OUTPUT/3035",
#   pattern = "*.vrt",
#   full.names = TRUE)

# coe_list <- lapply(coe_files, function(x){raster(x)})
# # LPD had different resolution.. Fede remade it

# Still, rasters are not aligned!!!!

# coe_sum <- rast(coe_files[grepl(names(band_order)[1], coe_files)])

if(exists("coe_stack")) {rm(coe_stack)}

tryCatch(
  coe_stack <- rast(paste0("../CoE_Europe_OUTPUT/3035_1221/",names(band_order), ".tif"))
)
if (!exists("coe_stack")) {
  print("data harmonization needed")
  ### sort out extent etc.
  # bypass by creating vrt
  file.create("./temp/buildvrtInputFiles.txt")
  write(paste0("../CoE_Europe_OUTPUT/3035_1221/",names(band_order), ".tif"),"./temp/buildvrtInputFiles.txt")
  # set target extent to something reasonable, e.g. that of zones_eu27
  expres <- paste("gdalbuildvrt -resolution user -te", paste(as.vector(round(ext(zones_eu27)[c(1,3,2,4)])), collapse = " "), "-tr 1000 1000 -separate -r nearest -input_file_list ../CoE_Europe_OUTPUT/3035_1221/buildvrtInputFiles.txt ../CoE_Europe_OUTPUT/3035_1221/coe_stack.vrt")
  system(expres)
  coe_stack <- writeRaster(rast("../CoE_Europe_OUTPUT/3035_1221/coe_stack.vrt"),
                           datatype = "INT1U",
                           filename = "./temp/coe_stack.tif")
  }

# different extents! Not aligned!
# shift in vrt introduced by fires.

# print(coe_stack)

# set band names.
names(coe_stack) <- names(band_order)

# get resolution
reso = res(coe_stack)
# multiplication factor to get area in sqkm from cell count
mf <- reso[1] * reso[2] * 1e-6

# rasterize study area : needed to couple zones with masks 
system.time(
rzones_eu27 <- zones_eu27 %>% 
  rasterize(coe_stack, field = "ID", filename = "./temp/rzones_eu27.tif")
)
# user  system elapsed 
# 0.875   0.135   1.692 

```

## Zonal statistics
Within each administrative unit, we summarize the information:
- area affected by a co-occurrence of issues (! need to identify which band is the sum of issues)
- area affected by each issue

We look at this in different land cover classes and in function of land productivity dynamics.

```{r zstat}
# we add LC, lpd and aridity to zones
# Land cover reclassied as 6 classes as used for stratification
LC <- rast("../CoE_Europe_OUTPUT/LC/LandCover_3035.tif") %>%
  # resample to match zones
  project(rzones_eu27, method = "near", filename = "./temp/LC.tif")

# change in LPD between reference period (2000-2015) and reporting (2004-2019)
LPD <- rast("../CoE_Europe_OUTPUT/LPD_sumNDVI/lpd_change_option_4.tif") %>%
  # reproject
  project(rzones_eu27, method = "near", filename = "./temp/LPD.tif")
# takes much more time than LC... possibly difference in encoding?

# aridity: problem: doesn't cover all land; would introduce no data
# so i drop it from the analysis

# sum raster
comb <- function(x, y, z){x + y * 1e4 + z * 1e5}
combined_zones <- lapp(sds(rzones_eu27, LC, LPD), 
             fun = comb,
             filename = "./temp/combined_zones.tif")
soi_zones <- lapp(sds(combined_zones, coe_stack[[1]]),
                 fun = function(x,y) {x + y *1e6},
                 filename = "./temp/soi_zones.tif")

# Issues
# zonal sum: sum of cells is equivalent to area in sq.km because we are in an equal area projection with 1000 m resolution AND data are 1 or 0
system.time(
stat_issues <- terra::zonal(coe_stack[[-1]], combined_zones, fun = "sum", na.rm = TRUE)
)
# i would like to have a way to know how much data
system.time(
stat_issues_data <- terra::zonal(coe_stack[[-1]], combined_zones, fun = function(x, ...){sum(!is.na(x))})
)
#   user  system elapsed 
# 99.005  11.661 115.369 
stat_issues_data %>% summarise(across(lpd_coe1000:wind, sum))
# only lpd has 6 pixels less. All the others have 4102196 data values.

# Sum of issues
system.time(
stat_soi <- terra::freq(soi_zones) %>% as.data.frame()
)
sum(stat_soi$count)
# 4102190, coming from the LPD.
# so we can safely use the areas from soi to analyse the individual issues.

# Prepare data for further analysis
issues_areas <- stat_issues %>%
  mutate(LPD = floor(ID * 1e-5)) %>%
  mutate(LC = floor((ID - LPD*1e5)*1e-4)) %>%
  mutate(ID = ID - LPD*1e5 - LC*1e4) %>%
  left_join(as.data.frame(zones_eu27) %>% dplyr::select(ID, NUTS_ID, CNTR_CODE, URBN_TYPE, MOUNT_TYPE, COAST_TYPE))
# I'm missing the total cell count with data at this stage...

soi_areas <- stat_soi %>%
  mutate(soi = floor(value * 1e-6)) %>%
  mutate(LPD = floor((value - soi*1e6) * 1e-5)) %>%
  mutate(LC = floor((value - soi*1e6 - LPD*1e5 )*1e-4)) %>%
  mutate(ID = value - soi*1e6 - LPD*1e5 - LC*1e4) %>%
  left_join(as.data.frame(zones_eu27) %>% dplyr::select(ID, NUTS_ID, CNTR_CODE, URBN_TYPE, MOUNT_TYPE, COAST_TYPE))

```
We visualize the tabular data as bar charts. Data are available at all NUTS levels.

```{r barplottotal}
# sum of issues
soi_eu <- soi_areas %>%
  group_by(soi) %>%
  summarise(area = sum(count))

p <- ggplot(soi_eu, aes(x = soi, y = area)) + geom_col()
p
# more than half of the (continental) EU has at least 3 coinciding issues.
```
More than half of the (continental) EU has at least 3 coinciding issues.

```{r soicntr}
soi_cntr <- soi_areas %>%
  group_by(soi, CNTR_CODE) %>%
  summarise(area = sum(count)) %>%
  pivot_wider(names_from = soi, values_from = area, values_fill = 0) %>%
  # sort countries alphabetically
  arrange(CNTR_CODE) %>%
  # add EU27 total
  bind_rows(soi_eu %>% mutate(CNTR_CODE = "EU27") %>% pivot_wider(names_from = soi, values_from = area, values_fill = 0))
knitr::kable(soi_cntr, cpation = "Areas of each EU country affected by a given number of coinciding issues.")
```
Countries do present a different distribution of the number of coinciding issues. 

```{r soicntrplot}
tmp <- soi_cntr %>%
  # compute total mapped area by country
  rowwise() %>%
  mutate(cntr_area = sum(c_across(`0`:`12`))) 
areas_cntr <- tmp %>% 
  dplyr::select(CNTR_CODE, cntr_area)%>%
  ungroup()
soi_cntr_pc <- tmp %>%
  mutate(across(`0`:`12`, ~round(.x/cntr_area*100))) %>%
  ungroup() %>%
  dplyr::select(-cntr_area) %>%
  # cntr_code to factor to fix order
  mutate(CNTR_CODE = factor(CNTR_CODE, levels = pull(tmp, CNTR_CODE))) %>%
  # to long table
  pivot_longer(cols = `0`:`12`, names_to = "soi", values_to = "area")
ggplot(soi_cntr_pc,
       aes(x = as.numeric(soi), y = area)) +
  geom_col() +
  facet_wrap(~CNTR_CODE, ncol = 7)
```
```{r soicntrmean}
soi_cntr_m <- soi_cntr_pc %>%
  mutate(soixarea = as.numeric(soi) * area) %>%
  group_by(CNTR_CODE) %>%
  arrange(as.numeric(soi), by_group = TRUE) %>%
  # mutating expressions are computed within groups
  mutate(cumarea = cumsum(area)) %>%
  summarise(mean = sum(soixarea)/100, 
            mode = soi[area == max(area)][1], 
            q1 = soi[cumarea == min(cumarea[cumarea>=25])],
            median = soi[cumarea == min(cumarea[cumarea>=50])],
            q3 = soi[cumarea == min(cumarea[cumarea>=75])]) %>%
  ungroup() %>% 
  mutate(across(mode:q3, as.numeric))
```

```{r soicntrmplot}
lim_mean <- soi_cntr_m %>% arrange(mean) %>% dplyr::select(CNTR_CODE) %>% unlist()
names(lim_mode) <-NULL
soicntrmplot <- ggplot(soi_cntr_m %>% mutate(CNTR_CODE = factor(CNTR_CODE, levels = lim_mean)), aes(x = CNTR_CODE)) +
  geom_pointrange(mapping = aes(y = median, ymin = q1, ymax = q3), colour = "red", show.legend = TRUE) +
  geom_point(mapping = aes(y = mean),show.legend = TRUE) +
  geom_point(mapping = aes(y = mode), shape = "x", size = 4, show.legend = TRUE) +
  ylab("# coinciding issues") + 
  xlab("Countries") 
  

soicntrmplot
```

We can distinguish groups of countries. Four countries have their peak of distribution (mode, x in the figure) at 1 single issue: LV, EE, FI, SE. Most of the mapped area of these countries have a single issue at a time, except for Sweden for which the median is 2. At the other extreme, six countries have more than half of their area subject to five or more coinciding issues: MT, DK, DE, LU, BE, NL. Half of the continental EU27 (without maritime territories) mapped in this analysis has at least three coinciding issues. There are at least two coinciding issues on 25% of the area, while another 25% of the area has at least four coinciding issues.

In the next section, we look at which are the most frequent issues.
```{r barplotissuestotal}
# sum of issues
issues_eu <- issues_areas %>%
  # to long table
  pivot_longer(lpd_coe1000:wind, names_to = "issue", values_to = "area") %>%
  group_by(issue) %>%
  summarise(area = sum(area)) %>%
  # sort by area
  arrange(area)

p <- ggplot(issues_eu, aes(x = factor(issue, levels = pull(issues_eu,issue)), y = area)) + 
  geom_col() +
  coord_flip()+ 
  xlab("") 
p
```

```{r }
issues_cntr <- issues_areas %>%
  group_by(CNTR_CODE) %>%
  summarise(across(lpd_coe1000:wind, sum)) %>%
  # sort countries alphabetically
  arrange(CNTR_CODE) %>%
  # add EU27 total
  bind_rows(issues_eu %>% mutate(CNTR_CODE = "EU27") %>% pivot_wider(names_from = issue, values_from = area, values_fill = 0)) %>%
  # rearrange columns
  dplyr::select(c("CNTR_CODE", pull(issues_eu,issue)))
knitr::kable(issues_cntr, cpation = "Areas of each EU country affected by a given issue.")

```

```{r issuescntrplot}
issues_cntr_pc <- issues_cntr %>%
  left_join(areas_cntr) %>%
  rowwise() %>%
  mutate(across(where(is.numeric), ~round(.x/cntr_area*100))) %>%
  ungroup() %>%
  dplyr::select(-cntr_area) %>%
  # cntr_code to factor to fix order
  mutate(CNTR_CODE = factor(CNTR_CODE, levels = pull(issues_cntr, CNTR_CODE))) %>%
  # to long table
  pivot_longer(fires:eutrof, names_to = "issue", values_to = "area") %>%
  # set issue as factor, with levels sorted by increasing area affected at EU scale
  mutate(issue = factor(issue, levels = pull(issues_eu,issue)))
  

ggplot(issues_cntr_pc,
       aes(x = issue, y = area)) +
  geom_col() +
  facet_wrap(~CNTR_CODE, ncol = 7) +
  coord_flip()
```
```

```{r}
###########

# soi <- tabs %>% 
#   # select sum of issues
#   dplyr::filter(grepl("coe", varname)) %>%
#   # order factor
#   mutate(sum_of_issues = factor(as.numeric(as.character(Var1)), levels = seq.int(0,nlayers(coe_stack)-1))) %>%
#   # compute area as Freq x pixel area
#   mutate(area = Freq * mf) %>%
#   dplyr::select(-Var1, -Freq, -varname)
# Area <- soi$area %>% sum()
# p <- ggplot(soi, 
#     aes(y=area/Area*100, x=sum_of_issues)) + #, fill = name, colour=Var1)) +
#     geom_col( size=1, width = 0.7) +
#     #####
#     labs(
#       y = "Percentage of area",#bquote(atop('%'~of~.(lu),(.(format(Area*1e-6,digits=3))~km^{2}))), #sprintf("%% of %s (%.3f M sq.km)",lu,Area*1e-12),
#       x = "Number of coinciding issues") +
#     scale_x_discrete(limits = levels(soi$sum_of_issues))
# 
# # ggsave(filename = "sumCE.png",
# #       p+theme(text = element_text(size = 24)), 
# #       path = "./fig",
# #       width=15,
# #       height=10,
# #       units="cm")
# 
# # table output
# # - sum_of_issues
# # | nuts
# soit_nuts3 <- soi %>%
#   # sort by increasing sum_of_issues
#   arrange(sum_of_issues) %>%
#   # make table wider with number of coinciding issues as columns
#   pivot_wider(id_cols = name, names_from = sum_of_issues, names_prefix = "soi", values_from = area) %>%
#   # compute rowwise total area
#   rowwise() %>%
#   mutate(total = sum(c_across(where(is.numeric)), na.rm = TRUE)) %>%
#   ungroup() %>%
#   # add official area (from eurostat)
#   left_join(reg_area3, by = c("name" = "geo"))
#   # get area percentages
#   mutate(across(starts_with("soi"), function(x){if_else(is.na(x), 0, round(x/total*100))}, .names = "{.col}_pc") )
#   #
# #  nuts0
# soit_nuts0 <- soi  %>%
#   # sort by increasing sum_of_issues
#   arrange(sum_of_issues) %>%
#   # make table wider with number of coinciding issues as columns
#   pivot_wider(id_cols = name, names_from = sum_of_issues, names_prefix = "soi", values_from = area)%>%
#   # change names to nuts0
#   mutate(name = substr(name, start = 1, stop = 2)) %>%
#   # group and summarise by name
#   group_by(name) %>%
#   summarise(across(starts_with("soi"), ~ sum(.x, na.rm = TRUE))) %>%
#   # compute row-wise total area
#   rowwise() %>%
#   mutate(total = sum(c_across(where(is.numeric)), na.rm = TRUE)) %>%
#   ungroup() %>%
#   # get area percentages
#   mutate(across(starts_with("soi"), function(x){if_else(is.na(x), 0, round(x/total*100))}, .names = "{.col}_pc") )
#   #

```

We can also look at the data, under different perspectives.

### Highlight on arid and semi-arid areas

### Dominant land use/cover classes (end of period)

### Land productivity dynamics

### Conventional ag

### Organic ag

### Regenerative ag

trends.earth outputs