---
title: "CoE_Europe_analysis"
author: "M. Weynants"
date: "12/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(raster)
library(sf)
library(dplyr)
library(ggplot2)
wd <- getwd()
```

## Load CoE output data
Assuming all issues have been processed, so that they :
- have the same equal area geographical coordinate system
- have the same extent
- are all dummy variables with values of 0 and 1
- have the same definition for nodata

```{r CoE_load}
coe_files <- list.files( 
  path = "../CoE_Europe_OUTPUT/3035",
  pattern = "*.tif",
  full.names = TRUE)

coe_list <- lapply(coe_files, function(x){raster(x)})
# LPD has different resolution

coe_sum <- raster(coe_files[grepl("coe", coe_files)][1])
coe_stack <- brick(coe_files[grepl("coe", coe_files)][2])

coe_files <- list.files( 
  path = "../CoE_Europe_OUTPUT/3035",
  pattern = "*.vrt",
  full.names = TRUE)

tryCatch(
  coe_stack <- raster::stack(coe_files)
)
if (!exists("coe_stack")) { 
  print("data harmonization needed")
  ### sort out extent etc.
  }

print(coe_stack)

# set band names.
# get band order from vrt
band_order <-  c(
  "coe" = "Sum of issues",
  "LPD" = "Land Productivity Dynamics Change",
  "Virtual" = "",
  "builtup" = "Built-up",
  "bws" = "Water stress",
  "eutro" = "Eutrophication",
  "fires" = "Fires",
  "gwdecline" = "Groundwater decline",
  "hhi" = "Household income",
  "irrig" = "Irrigation under water stress",
  "nitrogen" = "Nitrogen",
  "pop" = "Population",
  "soilacid" = "Soil acidification",
  "soilero" = "Water erosion",
  "treeloss" = "Tree loss",
  "windero" = "Wind erosion"
)
names(coe_stack) <- names(band_order)[-3]
# check that crs is equal area


# get resolution
reso = res(coe_stack)
# multiplication factor to get area in sqkm from cell count
mf <- reso[1] * reso[2] * 1e-6
```
## Load zones
The outputs of the convergence of evidence are summarised at administrative (or other) level.
```{r zone_load}

# load NUTS level 3 polygons in ETRS89 from EUROSTAT gisco API
url_zones <- "https://gisco-services.ec.europa.eu/distribution/v2/nuts/geojson/NUTS_RG_01M_2021_3035_LEVL_3.geojson"
zones <- geojsonsf::geojson_sf(url_zones)
# cite as © EuroGeographics for the administrative boundaries
st_crs(zones) <- CRS(projargs = "+init=epsg:3035")

zone_test <- zones[100:101,]
# zone_test <- st_transform(zone_test, crs(coe_stack))

# rasterize zones ? maybe not needed with extract... 
# But needed to couple zones with masks 

```

## Zonal statistics
Within each administrative unit, we summarize the information:
- area affected by each issue
- area affected by a co-occurrence of issues (! need to identify which band is the sum of issues)

```{r zstat}
# general summary stat by zone: sum of area for each issue + for number of issues
# loop on bands
# loop on values in band (either 0;1 or sum of issues)
# loop on zones 
# zonal or extract

# extract raster cell values for each zone
zone_ext <- extract(
  # raster object
  coe_stack,
  # zones as multipolygon
  zone_test, 
  # extrcation method: values
  method = 'simple')
# returns a list of lists with length = number of zones, with lists of raster cell values for each zone
names(zone_ext) <- zone_test$id

# function to tabulate issue by zone, returning a data.frame
tabFunc <- function (indx, extr, zone, zname) {
  # contingency tables
  ltb <- list()
  for (indb in seq(dim(extr[[1]])[2])) {
    dat <- as.data.frame(table(extr[[indx]][,indb]))
    dat$name <- zone[[zname]][[indx]]
    dat$varname <- dimnames(extr[[indx]])[[2]][indb]
    ltb[[indb]] <- dat
  }
  # list to data frame
  return(do.call("rbind", ltb))
}

# run on each zone. count raster cells for each unique value in coe_stack bands
tabs <- lapply(seq(zone_ext), tabFunc, zone_ext, zone_test, "NUTS_ID")
# returns a list of data frames. convert to data frame
tabs <- do.call("rbind", tabs)
# export result
if (!dir.exists("../temp/")){dir.create.dir("../temp/")}
write.csv(tabs, file = ("../temp/tabs_total.csv"))
```
We visualize the tabular data as bar charts. Data are available at all NUTS levels.

```{r barplottotal}
# sum of issues
soi <- tabs %>% 
  # select sum of issues
  dplyr::filter(varname == "coe") %>%
  # order factor
  mutate(sum_of_issues = Var1) %>%
  # compute area as Freq x pixel area
  mutate(area = Freq * mf)
Area <- soi$area %>% sum()
p <- ggplot(soi, 
    aes(y=area/Area*100, x=Var1)) + #, fill = name, colour=Var1)) +
    geom_col( size=1, width = 0.7) +
    #####
    labs(
      #y = bquote(atop('%'~of~.(lu),(.(format(Area*1e-6,digits=3))~km^{2}))), #sprintf("%% of %s (%.3f M sq.km)",lu,Area*1e-12),
      x = "Number of issues") +
    scale_y_continuous(limits=c(0,max(40,max(sum_areas[,which(names(sum_areas) %in% field)])/Area*100))) +
    scale_colour_manual(values=plt) + 
    scale_fill_grey(start = 0.7, end = 0.4) +
    guides(colour = "none", fill = guide_legend('')) +
    coord_flip() +
    theme(
      axis.title.x = element_text(hjust=1),
      panel.border = element_rect(fill=NA, colour = "gray40"),
      plot.background = element_rect(fill = "transparent"),
      panel.background = element_rect(fill = "transparent"),
      panel.grid.major.x = element_line(colour = "black"),
      # panel.grid.minor.x = element_line(element_blank()),
      # axis.line.x = element_line(colour = "black"),
      # axis.line.y = element_line(colour = "black"),
      axis.ticks = element_blank(),
      aspect.ratio = 1/.8
    )
ggsave(filename = paste0(gsub("/","_",cont_adjective),'_',luShort,"_sumCE_dry.png"),
      p+theme(text = element_text(size = 24)), 
      path = paste0("../graphs/",continent, "/", zone),
      width=15,
      height=10,
      units="cm")
```

We can also look at the data, under different perspectives.

### Highlight on arid and semi-arid areas

### Dominant land use/cover classes (end of period)

### Land productivity dynamics

### Conventional ag

### Organic ag

### Regenerative ag

trends.earth outputs